Concurrent Priority Queues

### 2014-03-24 presentation/presentation.tex
Remove framebreaks from concepts section
PQ, SL acronyms

\begin{frame}[allowframebreaks]{Concepts and Definitions}
\begin{itemize}
\item Safety conditions: nothing bad has happened yet
    \begin{itemize}
    \item \emph{Linearizability}: operations appear to take effect at a single point in time, the linearization point
    \item \emph{Quiescent consistency}: in a period of quiescence, semantics equivalent to some sequential ordering
    \item And others, i.e. \emph{sequential consistency}, \emph{serializability}
    \end{itemize}

\item Liveness conditions: something good eventually happens
    \begin{itemize}
    \item \emph{Lock-freedom}: at least a single process makes progress at all times
    \item \emph{Wait-freedom}: every process finishes in a bounded number of steps
    \item Further liveness conditions: \emph{Starvation-freedom}, \emph{Deadlock-freedom}
    \end{itemize}

\item \emph{Disjoint-access parallelism}: how well a data structure handles concurrent use by multiple
      threads within disjoint areas
\item Synchronization primitives:
    \begin{itemize}
    \item \ac{CAS}, \ac{FAA}, \ac{FAO}, \ac{TAS}
    \item \ac{DCAS}, \ac{DCSS}
    \end{itemize}

\end{itemize}

@@

\begin{frame}{\nameref{sec:related}}
\begin{itemize}
\item Non-standard synchronization primitives
    \begin{itemize}
    \item \citeauthor{liu2012lock}: Array-based PQ with \lstinline|ExtractMany|
    \item \citeauthor{israeli1993efficient}: Wait-free PQ
    \end{itemize}

@@

\item PQs in distributed memory systems
    \begin{itemize}
    \item \citeauthor{sanders1998randomized}: Local PQs, \lstinline|Insert| at random processor
    \end{itemize}

@@

\begin{itemize}
\item Naive PQ parallelization: single global lock \textrightarrow ~ sequential bottleneck
\item A first improvement: fine-grained locking using a lock per node
\item \fullcite{hunt1996efficient}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item A global lock remains
\item Heap is statically allocated
\item Frequent complex heap reorganization
\item Disjoint-access breaks down at high traffic levels
\item Inherent PQ bottleneck at the minimal node
\item Benchmarks show only limited scalability up to a low thread count
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Modern concurrent PQs are mostly based on \citeauthor{pugh1990skip}'s SkipList
\item Probabilistic ordered search structure, insertions and deletions in expected $O(\log n)$ time
\item No reorganizations
\item Simple implementation
\item Excellent disjoint-access properties
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item First SkipList-based PQ
\item \fullcite{shavit2000skiplist}
\item Initially lock-based, lock-free variant published in 2008
\item \fullcite{herlihy2012art}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Items are considered in the list once inserted on bottom level
\item Again, insertions bottom-up and deletions top-down
\item Deletions are split
    \begin{itemize}
    \item Logical deletion sets a \lstinline|deleted| flag
    \item Physical deletion performs actual pointer manipulations
    \end{itemize}
\item \lstinline|DeleteMin| attempts to logically delete the head node.
     On success: delete physically \& return node. Otherwise, continue with
     next node.
\item \lstinline|Insert| is equivalent to the SL insertion
\end{itemize}
\end{frame}

>>>>

\begin{frame}{Concepts and Definitions}
\framesubtitle{Safety conditions: nothing bad has happened yet}

\begin{itemize}
\item \emph{Linearizability}: operations appear to take effect at a single point in time, the linearization point
\item \emph{Quiescent consistency}: in a period of quiescence, semantics equivalent to some sequential ordering
\item And others, i.e. \emph{sequential consistency}, \emph{serializability}
\end{itemize}
\end{frame}

\begin{frame}{Concepts and Definitions}
\framesubtitle{Liveness conditions: something good eventually happens}

\begin{itemize}
\item \emph{Lock-freedom}: at least a single process makes progress at all times
\item \emph{Wait-freedom}: every process finishes in a bounded number of steps
\item Further liveness conditions: \emph{Starvation-freedom}, \emph{Deadlock-freedom}
\end{itemize}
\end{frame}

\begin{frame}{Concepts and Definitions}
\framesubtitle{Miscellaneous}

\begin{itemize}
\item \emph{Disjoint-access parallelism}: how well a data structure handles concurrent use by multiple
      threads within disjoint areas
\item Synchronization primitives:
    \begin{itemize}
    \item \ac{CAS}, \ac{FAA}, \ac{FAO}, \ac{TAS}
    \item \ac{DCAS}, \ac{DCSS}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{\nameref{sec:related}}
\begin{itemize}
\item Non-standard synchronization primitives
    \begin{itemize}
    \item \citeauthor{liu2012lock}: Array-based \ac{PQ} with \lstinline|ExtractMany|
    \item \citeauthor{israeli1993efficient}: Wait-free \ac{PQ}
    \end{itemize}

@@

\item \acp{PQ} in distributed memory systems
    \begin{itemize}
    \item \citeauthor{sanders1998randomized}: Local \acp{PQ}, \lstinline|Insert| at random processor
    \end{itemize}

@@

\begin{itemize}
\item Naive \ac{PQ} parallelization: single global lock \textrightarrow ~ sequential bottleneck
\item A first improvement: fine-grained locking using a lock per node
\item \fullcite{hunt1996efficient}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item A global lock remains
\item Heap is statically allocated
\item Frequent complex heap reorganization
\item Disjoint-access breaks down at high traffic levels
\item Inherent \ac{PQ} bottleneck at the minimal node
\item Benchmarks show only limited scalability up to a low thread count
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Modern concurrent \acp{PQ} are mostly based on \citeauthor{pugh1990skip}'s \ac{SL}
\item Probabilistic ordered search structure, insertions and deletions in expected $O(\log n)$ time
\item No reorganizations
\item Simple implementation
\item Excellent disjoint-access properties
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item First SkipList-based \ac{PQ}
\item \fullcite{shavit2000skiplist}
\item Initially lock-based, lock-free variant published in 2008
\item \fullcite{herlihy2012art}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Items are considered in the list once inserted on bottom level
\item Again, insertions bottom-up and deletions top-down
\item Deletions are split
    \begin{itemize}
    \item Logical deletion sets a \lstinline|deleted| flag
    \item Physical deletion performs actual pointer manipulations
    \end{itemize}
\item \lstinline|DeleteMin| attempts to logically delete the head node.
     On success: delete physically \& return node. Otherwise, continue with
     next node.
\item \lstinline|Insert| is equivalent to the \ac{SL} insertion
\end{itemize}
\end{frame}
